---
title: "ETC3250: Data visualisation"
author: "Professor Di Cook, Econometrics and Business Statistics"
date: "Week 6, class 2"
output:
  beamer_presentation: 
    theme: Monash
header-includes:
- \usepackage[3D]{movie15}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```

```{r echo=FALSE}
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(lubridate)
library(scales)
library(readr)
library(ggmap)
library(HLMdiag)
library(RColorBrewer)
library(gridExtra)
library(dichromat)
library(xkcd)
library(maps)
library(ggmap)
library(htmltools)
library(rworldmap)
library(boot)
```

## Data - Pedestrian Sensor

```{r echo=FALSE}
pedestrian <- read_csv("../data/Pedestrian_Counts.csv",
      col_names=c("DateTime","SensorID","SensorName","Counts"), skip=1)
pedestrian$Date <- as.Date(substr(pedestrian$DateTime, 1, 11), format="%d-%b-%Y")
pedestrian$Year <- year(pedestrian$Date)
pedestrian$Month <- month(pedestrian$Date, label=TRUE, abbr=TRUE)
pedestrian$Day <- wday(pedestrian$Date, label=TRUE) # and re-order days
pedestrian$Day <- factor(pedestrian$Day, levels = levels(pedestrian$Day)[c(2:7, 1)])
pedestrian$Hour <- as.numeric(substr(pedestrian$DateTime, 13, 14))
pedestrian <- pedestrian %>% group_by(SensorID) %>% 
  mutate(Time=as.numeric(Date - min(Date))*24 + Hour)
pedestrian <- pedestrian %>% filter(Date >= as.Date("2013-01-01"))
glimpse(pedestrian)
```


## Plotting points

```{r eval=FALSE}
ggplot(data=pedestrian, aes(x=Hour, y=Counts)) + geom_point()
```

```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=2.5}
ggplot(data=filter(pedestrian, SensorID==1, 
                   Date==as.Date("2013-01-01")), 
       aes(x=Hour, y=Counts)) + geom_point() +
  xlab("Hour") + ylab("Count") 
```

## Your turn

How is the data mapped to graphical elements?

- data: _______
- aesthetics: _________
- geom: ________
- transformations: _________

## Adding lines

```{r eval=FALSE}
ggplot(data=pedestrian, aes(x=Hour, y=Counts)) + geom_line()
```

```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=2.5}
ggplot(data=filter(pedestrian, SensorID==1, 
                   Date==as.Date("2013-01-01")), 
       aes(x=Hour, y=Counts)) + geom_line() +
  xlab("Hour") + ylab("Count") 
```

## Your turn

How is the data mapped to graphical elements?

- data: _______
- aesthetics: _________
- geom: ________
- transformations: _________

## Multiple days

```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=2.5}
ggplot(data=filter(pedestrian, SensorID==1, 
                   Day=="Tues"), 
       aes(x=Hour, y=Counts)) + geom_point() +
  xlab("Hour") + ylab("Count") 
```

## Use lines

```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=2.5}
ggplot(data=filter(pedestrian, SensorID==1, 
                   Day=="Tues"), 
       aes(x=Hour, y=Counts)) + geom_line() +
  xlab("Hour") + ylab("Count") 
```

That's not what I wanted ....

## For each day grouped

`+ geom_line(aes(group=Date))`

````{r echo=FALSE, fig.align='center', fig.width=5, fig.height=2.5}
ggplot(data=filter(pedestrian, SensorID==1, 
                   Day=="Tues"), 
       aes(x=Hour, y=Counts)) + geom_line(aes(group=Date)) +
  xlab("Hour") + ylab("Count") 
```

## Which is better?

`+ geom_line(aes(group=Date))` vs `+ geom_boxplot()`

````{r echo=FALSE, fig.align='center', fig.width=4.5, fig.height=2}
p1 <- ggplot(data=filter(pedestrian, SensorID==1, 
                   Day=="Tues"), 
       aes(x=Hour, y=Counts)) + geom_line(aes(group=Date)) +
  xlab("Hour") + ylab("Count") 
p2 <- ggplot(data=filter(pedestrian, SensorID==1, 
                   Day=="Tues"), 
       aes(x=factor(Hour), y=Counts)) + geom_boxplot() +
  xlab("Hour") + ylab("Count") + 
  scale_x_discrete(labels=c("0","","","","","5","","","","","10",
                            "","","","","15","","","","","20",
                            "","","",""))
grid.arrange(p1, p2, ncol=2)
```

## New example - Flying etiquette

[41% Of Fliers Think You’re Rude If You Recline Your Seat](http://fivethirtyeight.com/datalab/airplane-etiquette-recline-seat/)

```{r echo=FALSE}
fly <- read_csv("../data/flying-etiquette.csv")
glimpse(fly)
```

## Variables

- Mix of categorical and quantiative variables. 
- What mappings are appropriate? 
- Area for counts of categories, side-by-side boxplots for mixed pair. 

## Support

````{r echo=FALSE, fig.align='center', fig.width=4.5, fig.height=2.5}
ggplot(fly, aes(x=`How often do you travel by plane?`)) + 
  geom_bar() + coord_flip()
```

Categories are not sorted

## Sorted categories

```{r echo=FALSE, fig.align='center', fig.width=4.5, fig.height=2.5}
fly$`How often do you travel by plane?` <- 
  factor(fly$`How often do you travel by plane?`, levels=c(
    "Never","Once a year or less","Once a month or less",
    "A few times per month","A few times per week","Every day"))
ggplot(fly, aes(x=`How often do you travel by plane?`)) + geom_bar() + coord_flip()
```

## Filter data

````{r}
fly_sub <- fly %>% 
  filter(`How often do you travel by plane?` %in% 
    c("Once a year or less","Once a month or less")) %>%
  filter(!is.na(`Do you ever recline your seat when you fly?`)) %>%
  filter(!is.na(Age)) %>% filter(!is.na(Gender))
```

## Recline by height

````{r echo=FALSE, fig.align='center', fig.width=4, fig.height=2.5}
fly_sub$`Do you ever recline your seat when you fly?` <- factor(
  fly_sub$`Do you ever recline your seat when you fly?`, levels=c(
    "Never","Once in a while","About half the time",
    "Usually","Always"))
ggplot(fly_sub, aes(y=`How tall are you?`, x=`Do you ever recline your seat when you fly?`)) + geom_boxplot() + coord_flip()
```

## Cheat sheet

Take a look at the ggplot2 [Cheat sheet](http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf)

## Your turn

How many geoms are available in ggplot2? What is `geom_rug`?

```{r echo=FALSE, eval=FALSE}
A lot of geoms!
  
geom_rug adds tick marks to the axis, particularly for a scatterplot, or density, indicating values of data on that variable. 
```


## Your turn

What is the difference between `colour` and `fill`?

```{r echo=FALSE, eval=FALSE}
# colour is for 0 or 1-dimensional elements, and fill is for area (2-d) geoms
```

## Your turn

What does `coord_fixed()` do? What is the difference between this and using `theme(aspect.ratio=...)` or `coord_equal()`?

```{r echo=FALSE, eval=FALSE}
coord_fixed operates on the raw data values, but theme(aspect_ratio=...) works on the plot dimensions
```

## Your turn

What are scales? How many numeric transformation scales are there?

```{r echo=FALSE, eval=FALSE}
scales do the transformation between data values and graphical element value. most often it is applied to position along x, y which is common, to log or sqrt, .. there are 3 numeric transformations. 
```

## Your turn

What are position adjustments? When would they be used?

```{r echo=FALSE, eval=FALSE}
positions shift the location some from original coordinates. most often used with bar charts to stack, or put side-by-side
```

## Your turn 

- What type of plot do you need to make to explore the relationship between 
`Do you ever recline your seat when you fly?` and `Is it rude to recline your seat on a plane?`

- Use your cheat sheet to work out how to make it.

```{r echo=FALSE, eval=FALSE}
ggplot(fly_sub, aes(x=`Do you ever recline your seat when you fly?`)) +
  geom_bar() + 
  facet_wrap(~`Is itrude to recline your seat on a plane?`, ncol=3) +
  coord_flip()
```

## Facets

```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=2.5}
ggplot(fly_sub, 
       aes(x=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar() + coord_flip() + facet_wrap(~Gender)
```

## Facets

```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=2.5}
fly_sub$Age <- factor(fly_sub$Age, levels=c("18-29","30-44","45-60","> 60"))
ggplot(fly_sub, aes(x=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar() + coord_flip() + facet_grid(Age~Gender)
```

## Color palettes - default

```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=2.5}
p <- ggplot(fly_sub, aes(x=`In general, is itrude to bring a baby on a plane?`,
                    fill=Gender)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5)
p
```

What do we learn?

## Color palettes - brewer

```{r echo=FALSE, fig.align='center', fig.width=5, fig.height=2.5}
p + scale_fill_brewer(palette="Dark2") 
```

## Color blind-proofing

```{r fig.show='hide'}
library(scales)
library(dichromat)
clrs <- hue_pal()(3)
p + theme(legend.position = "none")
clrs <- dichromat(hue_pal()(3))
p + scale_fill_manual("", values=clrs) + theme(legend.position = "none")
```

---

```{r echo=FALSE, fig.width=5, fig.height=2.5, fig.align='center'}
clrs <- hue_pal()(3)
p + theme(legend.position = "none")
```

---

```{r echo=FALSE, fig.width=5, fig.height=2.5, fig.align='center'}
clrs <- dichromat(hue_pal()(3))
p + scale_fill_manual("", values=clrs) + theme(legend.position = "none")
```

## Perceptual principles

- Hierarchy of mappings: (first) position along an axis - (last) color (Cleveland, 1984; Heer and Bostock, 2009)
- Pre-attentive: Some elements are noticed before you even realise it.
- Color: (pre-attentive) palettes - qualitative, sequential, diverging.
- Proximity: Place elements for primary comparison close together. 
- Change blindness: When focus is interrupted differences may not be noticed.

## Hierarchy of mappings

1. Position - common scale (BEST)
2. Position - nonaligned scale
3. Length, direction, angle
4. Area
5. Volume, curvature
6. Shading, color (WORST)

See my [blog post](http://visiphilia.org/2016/08/03/CM-hierarchy) for a re-do with crowd-sourcing Aug 3, 2016. 

## Pre-attentive

"A preattentive visual property is one which is processed in spatial memory without our conscious action. In essence it takes less than 500 milliseconds for the eye and the brain to process a preattentive property of any image.""

Source: [Interaction Design Foundation](https://www.interaction-design.org/literature/article/preattentive-visual-properties-and-how-to-use-them-in-information-visualization)

## Can you find the odd one out?

```{r echo=FALSE, fig.width=3, fig.align='center'}
df <- data.frame(x=runif(100), y=runif(100), cl=sample(c(rep("A", 1), rep("B", 99))))
ggplot(data=df, aes(x, y, shape=cl)) + theme_bw() + 
  geom_point() +
  theme(legend.position="None", aspect.ratio=1)
```

## Is it easier now?

```{r echo=FALSE, fig.width=3, fig.align='center'}
ggplot(data=df, aes(x, y, colour=cl)) + 
  geom_point() +
  theme_bw() + 
  theme(legend.position="None", aspect.ratio=1)
```

## Color palettes

- Qualitative: categorical variables
- Sequential: low to high numeric values
- Diverging: negative to positive values

---

```{r, echo=FALSE, fig.height=5, fig.width=6}
library(RColorBrewer)
display.brewer.all()
```


## Proximity

```{r fig.show='hide'}
ggplot(fly_sub, 
       aes(x=`In general, is itrude to bring a baby on a plane?`,
                    fill=Gender)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5) +
  scale_y_continuous(breaks=seq(0,1,0.25), 
                     labels=c("0","","0.5","","1")) + 
  xlab("Rude to bring a baby?") + 
  theme(legend.position="bottom")
```

With this arrangement we can see proportion of gender within each rudeness category, and compare these across age groups.  How could we arrange this differently?

##


```{r echo=FALSE, fig.width=4.5, fig.height=2.5, fig.align='center'}
ggplot(fly_sub, 
       aes(x=`In general, is itrude to bring a baby on a plane?`,
                    fill=Gender)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5) +
  scale_y_continuous(breaks=seq(0,1,0.25), 
                     labels=c("0","","0.5","","1")) + 
  xlab("Rude to bring a baby?") + 
  theme(legend.position="bottom")
```


## Proximity

```{r fig.show='hide'}
ggplot(fly_sub, aes(x=Gender,
   fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5) +
  scale_y_continuous(breaks=seq(0,1,0.25), 
                     labels=c("0","","0.5","","1")) + 
  scale_fill_discrete("baby?") + 
  theme(legend.position="bottom")
```

##

```{r echo=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(fly_sub, aes(x=Gender,
          fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5) +   scale_y_continuous(breaks=seq(0,1,0.25), 
                     labels=c("0","","0.5","","1")) + 
  scale_fill_discrete("baby?") + 
  theme(legend.position="bottom")
```

What is different about the comparison now?

## Another arrangement

```{r fig.show='hide'}
ggplot(fly_sub, aes(x=Age,
          fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + 
  scale_y_continuous(breaks=seq(0,1,0.25), 
                     labels=c("0","","0.5","","1")) + 
  scale_fill_discrete("baby?") + 
  facet_wrap(~Gender, ncol=5) + 
  theme(legend.position="bottom")
```

##

```{r echo=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(fly_sub, aes(x=Age,
          fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + 
  scale_y_continuous(breaks=seq(0,1,0.25), 
                     labels=c("0","","0.5","","1")) + 
  scale_fill_discrete("baby?") + 
  facet_wrap(~Gender, ncol=5) + theme(legend.position="bottom")
```

## Themes

The `ggthemes` package has many different styles for the plots. Other packages such as `xkcd`, `skittles`, `wes anderson`, `beyonce`, ....

See the [vignette](https://cran.r-project.org/web/packages/xkcd/vignettes/xkcd-intro.pdf) for instructions on installing the xkcd font. 

##

```{r eval=FALSE}
library(xkcd)
ggplot(fly_sub, aes(x=Gender,
   fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + 
  facet_wrap(~Age, ncol=5) +
  scale_y_continuous(breaks=seq(0,1,0.25), 
                     labels=c("0","","0.5","","1")) + 
  scale_fill_discrete("baby?") + 
  theme_xkcd() + theme(legend.position="bottom")
```

##

```{r echo=FALSE, fig.width=4.5, fig.height=2.5, fig.align='center'}
library(xkcd)
ggplot(fly_sub, aes(x=Gender,
                    fill=`In general, is itrude to bring a baby on a plane?`)) + 
  geom_bar(position="fill") + coord_flip() + facet_wrap(~Age, ncol=5) +
  scale_y_continuous(breaks=seq(0,1,0.25), 
                     labels=c("0","","0.5","","1")) + 
  scale_fill_discrete("baby?") + 
  theme_xkcd() + theme(legend.position="bottom")
```

## Resources

- [Cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf)
- [ggplot2: Elegant Graphics for Data Analysis, Hadley Wickham](http://ggplot2.org/book/), [web site](http://ggplot2.org)
- [R Graphics Cookbook, Winston Chang](http://www.cookbook-r.com/Graphs/)
- [Naomi Robbins, Creating More Effective Graphs](http://www.nbr-graphs.com)
- [Antony Unwin, Graphical Data Analysis with R](https://www.crcpress.com/Graphical-Data-Analysis-with-R/Unwin/9781498715232)

## Share and share alike

This work is licensed under the Creative Commons Attribution-Noncommercial 3.0 United States License. To view a copy of this license, visit http://creativecommons.org/licenses/by-nc/ 3.0/us/ or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.
