---
title: "Supervised Classification Quiz 1"
date: "SOLUTION"
output: pdf_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
options(digits=1)
library(knitr)
```

```{r cache=FALSE, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold', fig.height=2.5, fig.width=2.5}
library(ggplot2)
library(tourr)
library(dplyr)
library(MASS)
```

## Question 1

Below are the summary statistics for linoleic acid, in the olive oils data (m=mean, s=sd).

```{r echo=FALSE, fig.align='center', fig.width=3, fig.height=3}
data(olive)
olive$region <- factor(olive$region, labels=c("South", "Sardinia", "North"))
olive.sub <- subset(olive, region != "South")
olive.sub$region <- factor(olive.sub$region)
olive.lda <- lda(region~linoleic, data=olive.sub, 
                 prior=c(0.5, 0.5))
#olive.lda$means
olive_s <- olive.sub %>% group_by(region) %>%
  summarise(m=mean(linoleic), s=sd(linoleic))
kable(olive_s)
ggplot(data=olive.sub, aes(x=linoleic, fill=region,
                           color=region)) + 
      geom_density(alpha=I(0.5)) +
  geom_vline(xintercept=961.78) +
  theme_bw() + theme(legend.position="none", aspect.ratio=1)
```

a. Sketch the LDA boundary on the plot.
b. A new observation with linoleic acid equal to 1000 would be classified into which region? `North`

## Question 2

The pooled variance is the same as the variance across all groups. `FALSE`

## Question 3

If the prior probability for group 1 is higher than group 2, the LDA boundary is actually shifted towards group 1? `NO`

## Question 4

Compute the classification error from the tabulation of true class vs predicted class below. `Error rate is` `r 15/(15+98+136)`

```{r echo=FALSE}
x <- table(olive.sub$region, predict(olive.lda, olive.sub)$class)
colnames(x) <- c("Pred Sard", "Pred North")
kable(x)
```

