---
title: "Wrangling Quiz"
date: "September 6, 2016"
output: pdf_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  echo=FALSE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
library(knitr)
```

```{r}
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(lubridate)
library(scales)
library(readr)
library(ggmap)
library(HLMdiag)
library(RColorBrewer)
library(gridExtra)
library(dichromat)
library(xkcd)
library(maps)
library(ggmap)
library(htmltools)
library(rworldmap)
library(boot)
```

## Question 1

a. `Getting your data into tidy form allows for analysing it in many different ways, as opposed to messy data which may be structured to tackle a single purpose.`  TRUE or FALSE

b. `Tidy data is like Lego, Messy data is like playmobile.`  TRUE or FALSE

## Question 2

Match the dplyr verb to the action

```{r}
x <- data.frame(verb=c("Arrange", "Summarise", "Filter", "Select", "Mutate", "Ungroup"), action=c("Produce subsets", "Reorder the rows", "Create new variables", "Removing existing grouping", "Choose columns","Make summaries of subsets"))
kable(x)
```

## Question 3

In the plot below we are looking at the relationship between humidity and air temperature as measured by Pacific Ocen buoys monitoring El Nino. There are missing values coded as 10% below the minimum on both variables. Circle the cases that are missing on humidity but not on air temperature.

```{r fig.width=3, fig.height=3}
library(MissingDataGUI)
data(tao)
tao_nomiss <- tao 
tao_nomiss$humidity[is.na(tao_nomiss$humidity)] <-
  min(tao_nomiss$humidity, na.rm=T) - 0.1*diff(range(tao$humidity, na.rm=T))
tao_nomiss$air.temp[is.na(tao_nomiss$air.temp)] <-
  min(tao_nomiss$air.temp, na.rm=T) - 0.1*diff(range(tao$air.temp, na.rm=T))
ggplot(tao_nomiss, aes(x=humidity, y=air.temp)) + geom_point() + theme(aspect.ratio=1)
```
