---
title: "Vis Quiz 1"
date: "August 30, 2016"
output: pdf_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```

```{r echo=FALSE}
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(lubridate)
library(scales)
library(readr)
library(ggmap)
library(HLMdiag)
library(RColorBrewer)
library(gridExtra)
library(dichromat)
library(xkcd)
library(maps)
library(ggmap)
library(htmltools)
library(rworldmap)
library(boot)
```

## Question 1


```{r echo=FALSE, fig.align='center', fig.width=9, fig.height=4}
pedestrian <- read_csv("../data/Pedestrian_Counts.csv",
      col_names=c("DateTime","SensorID","SensorName","Counts"), skip=1)
pedestrian$Date <- as.Date(substr(pedestrian$DateTime, 1, 11), format="%d-%b-%Y")
pedestrian$Year <- year(pedestrian$Date)
pedestrian$Month <- month(pedestrian$Date, label=TRUE, abbr=TRUE)
pedestrian$Day <- wday(pedestrian$Date, label=TRUE) # and re-order days
pedestrian$Day <- factor(pedestrian$Day, levels = levels(pedestrian$Day)[c(2:7, 1)])
pedestrian$Hour <- as.numeric(substr(pedestrian$DateTime, 13, 14))
pedestrian <- pedestrian %>% group_by(SensorID) %>% 
  mutate(Time=as.numeric(Date - min(Date))*24 + Hour)
pedestrian <- pedestrian %>% filter(Date >= as.Date("2013-01-01"))
ggplot(data=filter(pedestrian, SensorID==1), aes(factor(Hour), Counts)) +   geom_boxplot() +
  xlab("Hour") + ylab("Count") + 
  xkcdaxis(c(0,24), c(0,6000)) + theme_xkcd()
```

a. Data: Melbourne pedestrian sensor counts, Bourke St Mall, 2015
b. Aesthetics: __________
c. Geom: __________
d. (Extra credit) Theme: ___________

## Question 2

- In the hierarchy of mappings position along a common scale yields the most accurate reading of the data values from a plot. TRUE or FALSE

## Question 3

Fill in the blank

- A ___________ colour scale should be used when both negative (low) and positive (high) values of a quantitative variable are important. 

## Question 4

Both of these map use colour to display different levels of a categorical variable, but only one actually maps the variable to the colour aesthetic. Which one is it? A or B

```{r echo=FALSE}
library(ggplot2)
library(reshape2)
french_sub <- french_fries[french_fries$time == 10,]
french_fries_long <- gather(french_fries, key = variable, value = rating, potato:painty)
french_fries_long_sub <- french_fries_long[
  french_fries_long$time == 10,]
```

A
```{r fig.width=4, fig.height=3}
ggplot(data = french_sub, aes(x="1_potato", y=potato)) +
  geom_boxplot(fill = I("red")) +
    geom_boxplot(aes(x = "2_buttery", y = buttery), fill = I("orange")) +
    geom_boxplot(aes(x = "3_grassy", y = grassy), fill = I("yellow")) +
    geom_boxplot(aes(x = "4_rancid", y = rancid), fill = I("green")) +
    geom_boxplot(aes(x = "5_painty", y = painty), fill = I("blue")) +
    xlab("variable") + ylab("rating") + ggtitle("A")
```

B

```{r fig.width=5, fig.height=3}
ggplot(data = french_fries_long_sub, 
       aes(x=variable, y=rating, fill = variable)) + geom_boxplot() + ggtitle("B")
```

