---
title: "Lab session 3"
author: "Rob J Hyndman"
date: "11 August 2015"
output: html_document
published: false
---

We will build a regression model using the `Boston` data set from the `MASS` package.

The aim is to predict `medv` (median value of owner-occupied homes in each town) using characteristics of each town.

Find the best model you can using the available predictors.

Consider dropping predictors, using transformations of predictors, adding interaction terms, using splines, etc. 

You can easily add main effects and interactions by "multiplying" two predictors like this:
```{r}
lm(medv ~ ptratio*age, data=Boston)
```

We will determine which model is "best" based on the cross-validation statistic to be discussed later in the week. It can be computed using the following function.

```{r}
CV <- function(object)
{
  cv <- mean((residuals(object)/(1 - hatvalues(object)))^2, na.rm = TRUE)
  return(cv)
}
```

You can apply this function to the output from the `lm()` function.

The CV statistic is roughly equivalent to the MSE calculated on a test set (even though it is calculated on the training set).

Hints:

  * Regression plots can help you decide if a variable has a nonlinear effect on the response variable.
  * P-values can suggest potential variables to drop, but remember that p-values are misleading when predictors are correlated.

Try to work systematically, and keep track of the various models you have tried to avoid wasting time by re-fitting old models.
 
#Assignment 3

For your final (best) model, submit an Rmd file with the following information.

 * A summary of your chosen model.
 * A plot of the predicted values against the actual values for all towns. 
 * An interpretation of any nonlinear and interaction terms in the model.

