---
title: "ETC3250 Lab 6"
author: "Di Cook"
date: "1 September 2015"
output: pdf_document
---


#Principal component analysis, and dimension reduction

We will run PCA on the multiple test scores for Australian 15 year olds PISA test scores.

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(GGally)
```

# Assignment 6

Turn in two items: a .Rmd document, and the output .pdf from running it. No need to include the R output and plots in your pdf, but the code should be in the Rmd file. 

## Task 1

Read in the PISA data. How many students were tested? How many variables are included in the data set? Read the data dictionary to find out what the variables named ST08Q01 PV1MACC  PV2MACC  PV3MACC PV4MACC  PV5MACC  PV1MACQ  PV2MACQ  PV3MACQ  PV4MACQ PV5MACQ PV1MACS  PV2MACS  PV3MACS PV4MACS  PV5MACS  PV1MACU  PV2MACU  PV3MACU  PV4MACU  PV5MACU  PV1MAPE  PV2MAPE  PV3MAPE  PV4MAPE  PV5MAPE  PV1MAPF  PV2MAPF PV3MAPF  PV4MAPF  PV5MAPF  PV1MAPI  PV2MAPI  PV3MAPI  PV4MAPI  PV5MAPI are. Write a couple of sentences  describing them.

```{r, echo=FALSE, message=FALSE, results='hide'}
oz <- read.csv("../data/PISA-oz.csv") # You might need to change the directory to where the data is located on your computer
dim(oz)
```

## Task 2

Compute a PCA on the variables PV1MACC through PV5MAPI. Make a scree plot, and examine the principal components for the first 4. What proportion of variation in the data is explained by the first principal component? Second, third and fourth? 

```{r, echo=FALSE, message=FALSE, results='hide', fig.show='hide'}
library(dplyr)
math <- select(oz, PV1MACC:PV5MAPI)
math_pca <- prcomp(math, scale=T, retx=T)
plot(math_pca, type="l")
options(digits=2)
math_pca$sdev
math_pca$rotation[,1:4]
```

## Task 3 

Compute the average for each student for each of the different types of math tasks. Based on the PCA explain why this would be a reasonable thing to do. Make a scatterplot matrix of the average scores.

```{r, echo=FALSE, message=FALSE, results='hide', fig.show='hide'}
math_ave <- data.frame(MACC=apply(math[,1:5], 1, mean), 
                       MACQ=apply(math[,6:10], 1, mean), 
                       MACS=apply(math[,11:15], 1, mean), 
                       MACU=apply(math[,16:20], 1, mean), 
                       MAPE=apply(math[,21:25], 1, mean), 
                       MAPF=apply(math[,26:30], 1, mean), 
                       MAPI=apply(math[,31:35], 1, mean)) 
library(ggplot2)
library(GGally)
ggscatmat(math_ave)
```

## Task 4

Compute the average overall math score for each student (this means averaging PV1MATH-PV5MATH). Make a side-by-side boxplot of these scores by gender. Is there a difference in math scores for girls and boys? Write a few sentences explaining what you learn. (Note that the full range of math scores is 0-1000.)

```{r, echo=FALSE, message=FALSE, results='hide', fig.show='hide'}
m_g <- data.frame(gender=oz$ST04Q01, math=apply(oz[,31:35], 1, mean))
qplot(gender, math, data=m_g, geom="boxplot")
```

## Task 5 

How many different schools were included in the survey? Compute the average math score (average the averages) and standard deviation for each school, and make an ordered dotplot (with bars indicating one standard deviation above and below the mean) of these averages. Write a couple of sentences that describe how math scores vary across schools.

```{r, echo=FALSE, message=FALSE, results='hide', fig.show='hide', warning=FALSE}
length(unique(oz$SCHOOLID))
m_g <- data.frame(sch=oz$SCHOOLID, math=apply(oz[,31:35], 1, mean))
mga <- summarise(group_by(m_g, sch), m=mean(math), s=sd(math))
mga$sch <- factor(mga$sch, levels=order(mga$m))
qplot(sch, m, data=mga) + 
  geom_linerange(aes(x=sch, ymin=m-s, ymax=m+s)) + coord_flip()
```
